An examination of the high-order dynamic interactions underlying multi-dimensional timeseries data

### 2018-10-16 paper/main.tex
finished pass through timecorr-specific methods; still need to describe evaluation metrics

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\linenumbers

@@

\subsection*{Higher-order correlations}
Given a timeseries of dynamic correlations (e.g., obtained using
Eqn.~\ref{eqn:timecorr}), higher-order correlations reflect the
dynamic correlations between columns of $\mathbf{Y}$.  Given unlimited
computing resources, one could use repeated applications of
Equation~\ref{eqn:timecorr} to estimate these higher-order
correlations (i.e., substituting in the previous output, $\mathbf{Y}$,
for the input, $\mathbf{X}$ in the equation).  However, because each
output $\mathbf{Y}$ has $O(F^2)$ columns relative to $F$ columns in
the input $\mathbf{X}$, the output of Equation~\ref{eqn:timecorr}
grows with the square of the number of repeated applications (total
cost of computing $n$\textsuperscript{th} order correlations is
$O(F^{2(n - 1)})$ for $n \in \mathcal{J}, n > 0$).  When $F$ or $n$ is large,
this approach quickly becomes intractable.

To make progress in computing $\mathbf{Y_{n+1}}$, we can approximate
$\mathbf{Y}_n$ by computing an $O(F)$-dimensional embedding of
$\mathbf{Y}_n$, termed $\hat{\mathbf{Y}}_n$, and then we can apply
Equation~\ref{eqn:timecorr} to $\hat{\mathbf{Y}}_n$ rather than
directly to $\mathbf{Y}_n$.  This enables us to maintain $O(n)$
scaling with respect to $n$, rather than exponential scaling via the
direct approach.

@@

@@

\textbf{JRM TODO: Need references for each of the below, plus citation
  of \href{https://github.com/aestrivex/bctpy/wiki}{bctpy toolbox} (source of measures listed below)}
\begin{itemize}
\item Betweenness centrality: Wei; Node Betweenness; edge betweenness
  centrality; edge betweeness wei; 
\item Shannon-entropy based diversity coefficient
\item Eigenvector centrality, pagerank centrality
\item flow coefficient (Honey et al. 2007, PNAS)
\item k-coreness centrality (directed, undirected)
\item within-module degree
\item participation coefficient (signed, unsigned)
\item subgraph centrality
\end{itemize}

Idea: for each timepoint, compute relevant measure for node.  Result:
a $T$ by $F$ matrix characterizing how the given measure for each node
over time.  We can use this as $\hat{\mathbf{Y}}_n$ and then apply
Equation~\ref{eqn:timecorr} to compute $\hat{\mathbf{Y}}_{n+1}.$

\textbf{JRM STOPPED HERE}

>>>>

\documentclass[english]{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{setspace}
\usepackage{apacite}
\usepackage{natbib}
\usepackage{pxfonts}
\usepackage[utf8]{inputenc}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage[left]{lineno}
\usepackage{soul}
\linenumbers

@@

\subsection*{Higher-order correlations}
Given a timeseries of dynamic correlations (e.g., obtained using
Eqn.~\ref{eqn:timecorr}), higher-order correlations reflect the
dynamic correlations between columns of $\mathbf{Y}$.  Given unlimited
computing resources, one could use repeated applications of
Equation~\ref{eqn:timecorr} to estimate these higher-order
correlations (i.e., substituting in the previous output, $\mathbf{Y}$,
for the input, $\mathbf{X}$ in the equation).  However, because each
output $\mathbf{Y}$ has $\mathcal{O}(F^2)$ columns relative to $F$ columns in
the input $\mathbf{X}$, the output of Equation~\ref{eqn:timecorr}
grows with the square of the number of repeated applications (total
cost of computing $n$\textsuperscript{th} order correlations is
$\mathcal{O}(F^{2(n - 1)})$ for $n \in \mathcal{J}, n > 0$).  When $F$ or $n$ is large,
this approach quickly becomes intractable.

To make progress in computing $\mathbf{Y_{n+1}}$, we can approximate
$\mathbf{Y}_n$ by computing an $\mathcal{O}(F)$-dimensional embedding of
$\mathbf{Y}_n$, termed $\hat{\mathbf{Y}}_n$, and then we can apply
Equation~\ref{eqn:timecorr} to $\hat{\mathbf{Y}}_n$ rather than
directly to $\mathbf{Y}_n$.  This enables us to maintain $\mathcal{O}(n)$
scaling with respect to $n$, rather than exponential scaling via the
direct approach.

@@

Applying dimensionality reduction algorithms to $\mathbf{Y}$ yields a
$\hat{\mathbf{Y}}$ whose columns reflect weighted combinations (or
nonlinear transformations) of the original columns of $\mathbf{Y}$.
This has two main consequences.  First, with each repeated
dimensionality reduction, the resulting $\hat{\mathbf{Y}}_n$ has lower
and lower fidelity (with respect to what the ``true'' $\mathbf{Y}_n$
might have looked like without using dimensionality reduction to
maintain scalability).  In other words, computing $\hat{\mathbf{Y}}_n$
is a lossy operation.  Second, whereas the columns of $\mathbf{Y}_n$
may be mapped directly onto pairs of columns of $\mathbf{Y}_{n-1}$,
that mapping either becomes less cleanly defined in
$\hat{\mathbf{Y}}_n$ due to the reweightings and/or nonlinear
transformations.

@@

Graph theoretic measures take as input a matrix of interactions (e.g.,
using the above notation, an $F \times F$ correlation matrix or
binarized correlation matrix reconstituded from a single timepoint's
row of $\mathbf{Y}$) and return as output a set of $F$ measures
describing how each node (feature) sits within that interactions
matrix with respect to the rest of the population.  Common measures
include betweeness centrality~\citep[the proportion of shortest paths
between each pair of nodes in the population that involves the given
node in question; e.g., ][]{Newm05, OpsaEtal10, Bart04, GeisEtal08,
  Free77}; diversity and dissimilarity~\citep[characterizations of how
differently connected a given node is from others in the population;
e.g., ][]{Rao82, Lin09, RicoSzei06}; Eigenvector centrality and
pagerank centrality~\citep[measures of how influential a given node is
within the broader network; e.g., ][]{Newm08, Bona07, LohmEtal10,
  HaluEtal13}; transfer entropy and flow coefficients~\citep[a measure
of how much information is flowing from a given node to other nodes in
the network; e.g., ][]{HoneEtal07, Schr00}; $k$-coreness
centrality~\citep[a measure of the connectivity of a node within its
local sub-graph; e.g., ][]{AlvaEtal05, ChriFowl10}; within-module
degree~\citep[a measure of how many connections a node has to its
close neighbors in the network; e.g., ][]{RubiSpor10}; participation
coefficient~\citep[a measure of the diversity of a node's connections
to different sub-graphs in the network; e.g., ][]{RubiSpor10}; and
sub-graph centrality~\citep[a measure of a node's participation in all
of the network's sub-graphs; e.g., ][]{EstrRodr05}.

As an alternative to the above dimensionality reduction approach to
embedding $\mathbf{Y}_n$ in a lower-dimensional space, but still
allowing for scalable explorations of higher-order structure in the
data, we also explore using the above graph theoretic measures as a
means of obtaining $\hat{\mathbf{Y}}_n$.  In particular: for a given
graph theoretic measure, $\eta: \mathcal{R}^{F \times F} \rightarrow
\mathcal{R}^F$, we can use $\eta$ to tranform each row of
$\mathbf{Y}_n$ in a way that characterizes the corresponding
graph-theoretic properties of each column.  Whereas the dimensionality
reduction approach to computing $\hat{\mathbf{Y}}_n$ is lossy,
the graph-theory approach is lossless.  However, whereas the
dimensionality reduction approach maintains ties (direct or indirect)
to the original activity patterns reflected in $\mathbf{Y}_{n-1}$, the
graph-theory approach does not.  Instead, the graph-theory
characterizes the nature and timecourse of each feature's
\textit{participation} in the network.
